\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-07 paper/main.tex
working on addressing reviewer comments:   - fixed a "bug" in quail (https://github.com/ContextLab/quail/releases/tag/v0.2.2) related to computing permutation-corrected clustering scores    - regenerated all figures using a quick run of the updated clustering score metric, using N = 10 iterations    - I'm now rerunning the "full" analysis using N = 500 iterations (as we used in the previous implementation) to generate the production-ready figures and stats     - added a new discusion section on theoretical implications related to feature variability     - added description to methods on tempor

@@

Several prior studies have examined how varying the richness or experiences, or
the extensive of encoding, can affect memory. Although specific details
differ~\citep{BoniEtal22}, in general these studies have found that richer and
more deeply or extensively encoded experiences are remembered
better~\citep{HargEtal12, Mada21, MeinEtal20}. Our findings help to elucidate
an additional factor that may contribute to these phenomenon. For example, our
finding that participants better remember ``feature rich'' lists (where words'
appearances are varied) than ``reduced'' lists (where words' appearances are
held constant) only when those feature rich lists are presented \textit{after}
reduced lists suggests that some factors that influence the richness or depth
of encoding may be relative, rather than absolute. In other words,
\textit{changes} in richness may be more important than the overall complexity
or numbers of features.

We did \textit{not} find any evidence that changing words' appearances
\textit{harmed} memory performance, e.g., by distracting them with irrelevant
information~\citep{MarsEtal15, Lang05, MarsEtal12, ReinEtal92}. Nor did we find
any evidence that \textit{changes} in the presence of potentially
``distracting'' features adversely affected memory. For example, when we
increased or decreased the variability in words' appearances on late versus
early lists (as in the reduced (early) and reduced (late) conditions), we found
no evidence that this harmed participants' memories.

% add something about multiple pathways to recall (for different feature dimensions)

>>>>

\subsubsection*{Factoring out the effects of temporal clustering}

For a given list of words, if the values along two feature dimensions (e.g.,
category and size) are correlated, then the clustering scores for those two
dimensions will also be correlated. When lists are sorted along a given feature
dimension, the sorted feature values will also tend to be correlated with the
serial positions of the words in the list. This means that the temporal
clustering score will \textit{also} tend to be correlated with the clustering
scores for the sorted feature dimension. These correlations mean that it can be
difficult to specifically identify when participants are using one feature
versus another (or a manipulated feature versus temporal information) to
organize or search their memories.

We developed a permutation-based procedure to factor out the effects of
temporal clustering from the clustering scores for each feature dimension. For
a given set of recalled items (whose presentation positions are given by $x_1,
x_2, x_3, …, x_N$), we circularly shift the presentation positions by a
randomly chosen amount (between 1 and the list length) to obtain a new set of
items. Since the new set of items will have the same (average) temporal
distances between successive recalls, the temporal clustering score for the new
set of items is equal (on average) to the temporal clustering score for the
original recalls. However, we can then re-compute the feature clustering score
for those new items. Finally, we can compute a ``temporally corrected'' feature
clustering score by computing the average percentile rank of the observed (raw)
feature clustering score within the distribution of circularly shifted feature
clustering scores. This new temporally corrected score provides an estimate of
the observed degree of feature clustering over and above what could be
accounted for by temporal clustering alone.

While these temporally corrected clustering scores are useful for identifying
when feature clustering cannot be accounted for by temporal clustering alone,
they are \textit{not} necessarily valid estimates of the ``true'' degree to
which participants are organizing their memories along a given feature
dimension. For example, on a list where the presentation order and feature
values (along the given feature dimension) are perfectly correlated, the
temporally corrected score will have an expected value of 0.5 no matter which
words (or in what order) are recalled. Therefore these temporally corrected
clustering scores are interpretable only to the extent that presentation order
and feature values are decoupled.

@@

Several prior studies have examined how varying the richness or experiences, or
the extensive of encoding, can affect memory. Although specific details
differ~\citep{BoniEtal22}, in general these studies have found that richer and
more deeply or extensively encoded experiences are remembered
better~\citep{HargEtal12, Mada21, MeinEtal20}. Our findings help to elucidate
an additional factor that may contribute to these phenomenon. For example, our
finding that participants better remember ``feature rich'' lists (where words'
appearances are varied) than ``reduced'' lists (where words' appearances are
held constant) only when those feature rich lists are presented \textit{after}
reduced lists suggests that some factors that influence the richness or depth
of encoding may be relative, rather than absolute. In other words,
\textit{increases} in richness (e.g., relative to a recency-weighted baseline)
may be more important than the overall complexity or numbers of features.

Some prior studies have suggested that people can ``cue'' their memories using
different ``strategies'' or ``pathways'' for searching for the target
information. For example, modern accounts of free recall typically posit that
memory search typically begins by matching the current state of mental context
with the contexts associated with other items in memory~\citep{Kaha20}. Since
context is the defining hallmark of episodic memory~\citep{Tulv83},
context-based search can be described as an ``episodic'' pathway to recall.
When episodic cueing fails to elicit a match, participants may then search for
items that are similar to the current mental context or mental state along
other dimensions, such as semantic similarity~\citep{DavaEtal03, SochEtal09}.
These multiple pathways accounts of memory search also provide a potential
explanation of why participants might have an easier time remembering richer
stimuli (or experiences): richer stimuli and experiences might have more
features that could be used to cue memory search. Our work suggests that there
may be some additional factors at play with respect to the \textit{dynamics} of
these processes. In particular, we only observed memory benefits for ``richer''
stimuli when they were encountered after more ``impoverished'' stimuli (in the
reduced (early) condition). This suggests that the pathways available to recall
a given item may also depend on recent prior experiences.

We did \textit{not} find any evidence that changing words' appearances
\textit{harmed} memory performance, e.g., by distracting them with irrelevant
information~\citep{MarsEtal15, Lang05, MarsEtal12, ReinEtal92}. Nor did we find
any evidence that \textit{changes} in the presence of potentially
``distracting'' features adversely affected memory. For example, when we
increased or decreased the variability in words' appearances on late versus
early lists (as in the reduced (early) and reduced (late) conditions), we found
no evidence that this harmed participants' memories. One potential
interpretation under the ``multiple pathways to recall'' framework is that the
availability of multiple pathways to recall do not appear to specifically
interfere with each other.
